Hydra battles in Coq

### 2020-09-29 doc/hydras.tex
correcting typos

Let us comment some features of this table:

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals that  follow dots immediately (displayed in red) are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item All the ordinals shown in this list are either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\subsection{Well-ordered sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness] $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. Please refer to any documenation on \coq{} for having more details on well-foundedness and accessibility.}.

@@

The best known examples of well-ordered sets are the set $\mathbb{N}$ of natural numbers (with the usual order $<$), as well as any finite segment $[0,i)=\{j\in\mathbb{N}|j<i\}$.
The disjoint union of two copies of $\mathbb{N}$, \emph{i.e.} the set $\{0,1\}\times\mathbb{N}$ is also well-ordered,
with respect to the order below:

@@

In a set-theoretic framework, one can consider any ordinal $\alpha$ as a well-ordered set, whose  elements are just the ordinals strictly less than $\alpha$, \emph{i.e.} the \emph{segment} $\mathbb{O}_\alpha=[0, \alpha)$. So, one can speak about \emph{finite}, \emph{infinite}, \emph{countable}, etc., ordinals. Nevertheless, since we work with type theory, 
we do not identify ordinals as sets of ordinals, but the correspondance between ordinals and sets of ordinals is the function that maps $\alpha$ to $\mathbb{O}_\alpha$.
For instance $\mathbb{O}_\omega=\mathbb{N}$, and $\mathbb{O}_7=\{0,1,2,3,4,5,6\}$.

@@

\begin{itemize}
\item A ``mathematical'' representation of the set of countable ordinal numbers, afer Kurt Schütte~\cite{schutte}. This representation uses several (hopefully harmess) axioms. It is used as a reference for proving the correctness of ordinal notations.
\item A family of \emph{ordinal notations}, \emph{i.e.} data types used to represent segments $[0,\mu)$, where $\mu$ is some countable ordinal. Each ordinal notation is defined inside the Calculus of Inductive Constructions (without axioms). Many functions are defined, allowing proofs by computation. Note that proofs of 
correctness of a given ordinal notation with respect to Schütte's model use axioms.
Please execute the \texttt{Print Assumptions} command in case of doubt.
\end{itemize}

@@

Chapter~\ref{chap:schutte} of this document presents an adaptation to \coq{} of an axiomatization in classical logic of the set of countable ordinals by K. Schütte~\cite{schutte}. 
That formalization is quite complex and technical and unshamedly non-constructive,  so we put its description  in the last chapter of this document.

@@

In our adaptation to \coq{}, we declare a type \texttt{Ord}, a binary relation \texttt{lt} (with infix notation \texttt{``\_<\_''}, and assume Schütte's axiom. In Chapter~\ref{chap:schutte},
we derive some interesting properties of countable ordinals from these axioms.

@@

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call ordinal notation for $\alpha$ a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A function for comparing two ordinals, whose correctnes implies that the reflexive closure of $<$ is a \emph{total order}.
\end{itemize}

@@

The following class definition, parameterized with a type $A$, a binary relation \texttt{lt} on $A$, specifies that \texttt{lt} is a well-founded strict order. The reflexive closure of \texttt{lt}, (called \texttt{le}, for ``less  or equal than'') is a total decidable order, implemented through a comparison function \texttt{compare}.

@@

Lemma wf_measure  {A:Type}(lt: relation A)
            {compare : A -> A -> comparison}
            {on : ON lt compare}
            {B : Type}
            (m : B -> A) :
  well_founded (measure_lt m). 
\end{Coqsrc}

@@

 We consider a new strict order
on the disjoint sum of the associated types, by putting all elements of \texttt{A} before the elements of \texttt{B} (thanks to Standard Library's relation operator \texttt{le\_AsB}).
(please look at Figure~\ref{fig:omega-plus-omega} for an example).

@@

\begin{Coqsrc}
Example ex1' : fin 7 o< omega.
Proof. constructor. Qed.
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition succb (alpha: t) := match alpha with
                                   inr (S  _) | inl (S _) => true
                                 | _ => false
                                 end.

@@

\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.Generic.html}{OrdinalNotations.Generic}}
\begin{Coqsrc}
  Definition ZeroLimitSucc_dec {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  forall alpha,
    {Least alpha} +
    {Limit alpha} +
    {beta: A | Successor alpha beta}.
\end{Coqsrc}

@@

In order to build an ordinal notation out the type \texttt{t}, we have to define a comparison function, and prove that the order \texttt{lt} is well-founded.

\begin{Coqsrc}
Definition compare (alpha beta: t) : comparison :=
  match Nat.compare (fst alpha) (fst beta) with
    Eq => Nat.compare (snd alpha) (snd beta)
  | c => c
  end.

@@

Instance Omega2 : ON  lt compare.
\end{Coqsrc}

@@

Using the lemma of Sect.~\vref{sect:measure-ON}, we can define easily the function which merges two lists.

@@

In Sect.~\vref{omega-case}, we proved that there exists no variant towards \texttt{nat}
(\emph{i.e.} the ordinal $\omega$) for proving the termination of all hydra battles.
We  prove now that even the ordinal $\omega^2$ is also insufficient for this purpose.

@@

The proof of the inequality \texttt{m big\_h o<= m small\_h} is quite more complex than in Sect~\ref{omega-case}.  If we consider some ordinal $\alpha=(i,j)$, where $i>0$, there exists an infinite number of
ordinals stricly less than $\alpha$, and there exists an infinite number of battles that start from
$\iota(\alpha)$. In effect, at any configuration $\iota(k,0)$, where $k>0$, the hydra can freely chose any replication number. Intuitively, the measure of such a hydra must be large enough for taking into account
all the possible battles issued from that hydra.
Let us give more technical details.

@@

Thus, starting from any inequality $q < p$ on type \texttt{t}, we can build 
by transfinite induction over \texttt{p} a battle 
that transforms the hydra $\iota(p)$ into $\iota(q)$.

@@

Since $<$ is a strict order (irreflexive  and transitive) on \texttt{nat*nat}, we can conclude that there is no
variant for termination on the lexicographic square of $(\mathbb{N},<)$.

@@

\section{Comparing Ordinal Notations}

Finally, it may be interesting to compare an ordinal notation with the more theoretical model from Schütte (well, at least with our formalization of that model). This would be a relative proof of correctenss of our ordinal  notations.

For instance, the following theorem tells that \texttt{Epsilon0}, our notation system for the segment $[0,\epsilon0)$ is correct with respect to the countable ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

Finally, it is interesting to show that, if $i<j$, then the segment $[0,i)$ is a ``sub-segment'' of
$[0,j)$. Since the terms  ($t\;i$) and ($t\;j$) are not convertible, we consider a ``cast'' 
function $\iota$ from ($t\;i$) into ($t\;j$), and prove that this function is  a monotonous bijection  from ($t\;i$) to
the segment $[0,i)$ of ($t\;j$).

@@

\paragraph{Related work}
Finite ordinals are also formalized in MathComp~\cite{SSR}.  See also Adam Chlipala's \emph{CPDT}~\cite{chlipalacpdt2011} for a thorough study of the use of dependent types.

@@

In some cases we want to show that two notation systems describe the same segment (for instance $[0,3+\omega)$ and $[0,\omega($). For this purpose, one may prove that the two notation systems are order-isomorphic.

@@

{\it \textbf{Note:} This property reflects the equality $i+\omega=\omega$ we will prove in larger notation systems, as well as in Schütte's model}

@@

\textbf{Hint:}   Beware of the instances associated with sum and product of notations !
  You may consider additional fields such as ``there is a greatest element'' 
to make the sum and product of notations ``compositional''.

@@

Theorem embedding : 
     fun_bijection (nf: Ensemble T1)  (members epsilon0) inject.
 \end{Coqsrc}
\subsection{Remarks}
Let us recall that the library \href{../src/html/hydras.Schutte.Schutte.html}%
{\texttt{Schutte}} depends on five \emph{axioms} and lies explicitly in the  
framework of classical logic with a weak version of the axiom of choice
(please look at the documentation of
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Logic.ChoiceFacts.html}{\texttt{Coq.Logic.ChoiceFacts}}).
Nevertheless, the other modules:
\href{../src/html/hydras.Epsilon0.Epsilon0.html}%
{\texttt{Epsilon0}},
\href{../src/html/hydras.Hydra.Hydra.html}%
{\texttt{Hydra}}, et 
\href{../src/html/hydras.Gamma0.Gamma0.html}%
{\texttt{Gamma0}}
do not import any axioms and are really constructive.

>>>>

Let us comment some features of this figure:

\begin{itemize}
\item The ordinals are listed in a strictly increasing order. 
\item Dots : ``$\ldots$'' stand for an infinite sequence of ordinals, not shown for lack of space. For instance, the ordinal $42$ is not shown in the first line, but it exists, between $17$ and $\omega$.
\item Each ordinal printed in black is the immediate successor of another ordinal. We call it a \emph{successor} ordinal. For instance, $12$ is the successor of $11$, and $\omega^4+1$ the successor of $\omega^4$.
\item Ordinals (displayed in red)  that  follow immediately dots are called \emph{limit ordinals}. With respect to the order induced by this sequence, any limit ordinal $\alpha$ is the least upper bound of  the set $\mathbb{O}_\alpha$ of all ordinals strictly less than $\alpha$.
\item
For instance $\omega$ is the least upper bound of the set of all finite ordinals (in the first line). It is also the first limit ordinal, and the first infinite ordinal, in the sense that 
the set $\mathbb{O}_\omega$ is infinite.
\item The ordinal $\epsilon_0$ is the first number which is equal to its own exponential of base $\omega$. It plays an important role in proof theory, and is particularly studied in chapters~\ref{chap:T1} to \ref{chap:alpha-large}.
\item Any ordinal is  either the ordinal \textcolor{blue}{$0$},
a successor ordinal, or a \textcolor{red}{limit ordinal}.
\end{itemize}

@@

\subsection{Well-ordered sets}
Let us start with some definitions.
A  \emph{well-ordered set} is a set provided with a binary relation $<$ which has the following properties.
\begin{description}
\item[irreflexivity] : $\forall x\in A, x\not< x$
\item[transitivity] : $\forall x\,y\,z\in A, x<y \Rightarrow y<z \Rightarrow x<z$
\item[trichotomy]: $\forall x\,y\in A, x<y \vee x = y \vee y < x$
\item[well foundedness]: $<$ is well-founded (every element of $A$ is accessible)\footnote{In classical mathematics, we would say that there is no infinite sequence $a_1>a_2> \dots a_n> a_{n+1}\dots$ in $A$. Please refer to any documenation on \coq{} for having more details on well-foundedness and accessibility.}.

@@

The best known examples of well-ordered sets are the set $\mathbb{N}$ of natural numbers (with the usual order $<$), as well as any finite segment $[0,i)=\{j\in\mathbb{N}\,|\,j<i\}$.
The disjoint union of two copies of $\mathbb{N}$, \emph{i.e.} the set $\{0,1\}\times\mathbb{N}$ is also well-ordered,
with respect to the order below:

@@

In a set-theoretic framework, one can consider any ordinal $\alpha$ as a well-ordered set, whose  elements are just the ordinals strictly less than $\alpha$, \emph{i.e.} the \emph{segment} $\mathbb{O}_\alpha=[0, \alpha)$. So, one can speak about \emph{finite}, \emph{infinite}, \emph{countable}, etc., ordinals. Nevertheless, since we work within type theory, 
we do not identify ordinals as sets of ordinals, but the correspondance between ordinals and sets of ordinals is the function that maps $\alpha$ to $\mathbb{O}_\alpha$.
For instance $\mathbb{O}_\omega=\mathbb{N}$, and $\mathbb{O}_7=\{0,1,2,3,4,5,6\}$.

@@

\begin{itemize}
\item A ``mathematical'' representation of the set of countable ordinal numbers, afer Kurt Schütte~\cite{schutte}. This representation uses several (hopefully harmless) axioms. We use it as a reference for proving the correctness of ordinal notations.
\item A family of \emph{ordinal notations}, \emph{i.e.} data types used to represent segments $[0,\mu)$, where $\mu$ is some countable ordinal. Each ordinal notation is defined inside the Calculus of Inductive Constructions (without axioms). Many functions are defined, allowing proofs by computation. Note that proofs of 
correctness of a given ordinal notation with respect to Schütte's model obviously use axioms.
Please execute the \texttt{Print Assumptions} command in case of doubt.
\end{itemize}

@@

Chapter~\ref{chap:schutte} of this document presents an adaptation to \coq{} of an axiomatization in classical logic of the set of countable ordinals by K. Schütte~\cite{schutte}. 
That formalization is quite complex, technical and unshamedly non-constructive,  so we put its description  in the last chapter of this document.

@@

In our adaptation to \coq{}, we declare a type \texttt{Ord}, a binary relation \texttt{lt} (with infix notation \texttt{"\_<\_"}, and assume Schütte's axiom. In Chapter~\ref{chap:schutte},
we derive some interesting properties of countable ordinals from these axioms.
It is interesting to compare proofs of a given property (for instance the associativity of addition) in the computatuinal framework of some ordinal notation, and in the axiomatic model of Schütte.

@@

Let $\alpha$ be some (countable) ordinal; 
in \coq{} terms, we call ordinal notation for $\alpha$ a structure composed 
of:
\begin{itemize}
\item A data type $A$ for representing all ordinals strictly below $\alpha$,
\item A well founded order $<$ on $A$, 
\item A correct function for comparing two ordinals. Note  that the reflexive closure of $<$ is thus a \emph{total order}.
\end{itemize}

@@

The following class definition, parameterized with a type $A$, a binary relation \texttt{lt} on $A$, specifies that \texttt{lt} is a well-founded strict order. The reflexive closure of \texttt{lt}, (called \texttt{le}, for ``less  or equal than'') is a total decidable order, implemented through a comparison function \texttt{compare}.  The correctness of this function is expressed through Stdlib's type 
\texttt{Datatypes.CompareSpec}.

\begin{Coqsrc}
Inductive CompareSpec (Peq Plt Pgt : Prop) : comparison -> Prop :=
    CompEq : Peq -> CompareSpec Peq Plt Pgt Eq
  | CompLt : Plt -> CompareSpec Peq Plt Pgt Lt
  | CompGt : Pgt -> CompareSpec Peq Plt Pgt Gt
\end{Coqsrc}

@@

Lemma wf_measure  {A:Type}(lt: relation A)
            {compare : A -> A -> comparison}
            {on : ON lt compare}
            {B : Type}
            (m : B -> A):  well_founded (measure_lt m). 
\end{Coqsrc}

@@

 We consider a new strict order
on the disjoint sum of the associated types, by putting all elements of \texttt{A} before the elements of \texttt{B} (thanks to Standard Library's relation operator \texttt{le\_AsB}).

@@

\begin{Coqsrc}
Example ex1' : fin 7 o< omega.
Proof. constructor. Qed.

Lemma lt_omega alpha : 
     alpha o< omega <-> exists n:nat,  alpha = fin n.
(* ... *)
\end{Coqsrc}

@@

\begin{Coqsrc}
Definition succb (alpha: t) : bool := match alpha with
                                 | inr (S  _) | inl (S _) => true
                                 | _ => false
                                 end.

@@

\noindent\emph{From Module~\href{../src/html/hydras.OrdinalNotations.Generic.html}{OrdinalNotations.Generic}}
\begin{Coqsrc}
  Definition ZeroLimitSucc_dec {A:Type}{lt: relation A}
           {compare : A -> A -> comparison}
           {on : ON lt compare} :=
  forall alpha,
    {Least alpha} +
    {Limit alpha} +
    {beta: A | Successor alpha beta}.
(* ... *)
\end{Coqsrc}

@@

In order to build an ordinal notation out of the type \texttt{t}, we have to define a comparison function, and prove that the order \texttt{lt} is well-founded.

\begin{Coqsrc}
Instance lt_strorder : StrictOrder lt.
(* ... *)

Definition compare (alpha beta: t) : comparison :=
  match Nat.compare (fst alpha) (fst beta) with
    Eq => Nat.compare (snd alpha) (snd beta)
  | c => c
  end.

@@

Lemma lt_wf : well_founded lt.

Instance Omega2 : ON  lt compare.
Proof.
  split.
  - apply lt_strorder.
  - apply lt_wf.
  - apply compare_correct.
Qed.
\end{Coqsrc}

@@

Using the lemma of Sect.~\vref{sect:measure-ON}, we can define easily a total function which merges two lists.

@@

In Sect.~\vref{omega-case}, we proved that there exists no variant towards \texttt{nat}
(\emph{i.e.} the ordinal $\omega$) for proving the termination of all hydra battles.
We  prove now that  the ordinal $\omega^2$ is also insufficient for this purpose.

@@

The proof of the inequality \texttt{m big\_h o<= m small\_h} is quite more complex than in Sect~\ref{omega-case}.  If we consider any ordinal $\alpha=(i,j)$, where $i>0$, there exists an infinite number of
ordinals stricly less than $\alpha$, and there exists an infinite number of battles that start from
$\iota(\alpha)$. Indeed, at any configuration $\iota(k,0)$, where $k>0$, the hydra can freely choose any replication number. Intuitively, the measure of such a hydra must be large enough for taking into account
all the possible battles issued from that hydra.
Let us now give more technical details.

@@

Thus, starting from any inequality $q < p$ on type \texttt{t}, we can build 
by \emph{transfinite induction} (\emph{i.e.} well-founded) over \texttt{p} a battle 
that transforms the hydra $\iota(p)$ into $\iota(q)$.

@@

Since $<$ is a strict order (irreflexive  and transitive) on \texttt{nat*nat}, we infer that there is no
variant for termination on the lexicographic square of $(\mathbb{N},<)$.

@@

@@

\begin{remark}
Finite ordinals are also formalized in MathComp~\cite{SSR}.  See also Adam Chlipala's \emph{CPDT}~\cite{chlipalacpdt2011} for a thorough study of the use of dependent types.  
\end{remark}

%%%

\section{Comparing two ordinal notations}

We can also show that, if $i<j$, then the segment $[0,i)$ is a ``sub-segment'' of
$[0,j)$. Since the terms  ($t\;i$) and ($t\;j$) are not convertible, we consider a ``cast'' 
function $\iota$ from ($t\;i$) into ($t\;j$), and prove that this function is  a monotonous bijection  from ($t\;i$) to
the segment $[0,i)$ of ($t\;j$).

@@

Finally, it may be interesting to compare an ordinal notation with the more theoretical model from Schütte (well, at least with our formalization of that model). This would be a relative proof of correctenss of the considered  ordinal  notation.

For instance, the following theorem tells that \texttt{Epsilon0}, our notation system for the segment $[0,\epsilon0)$ is a correct implementation of the theoretically defined  ordinal $\epsilon_0$
(see chapter~\ref{chap:schutte} for more details).

@@

In some cases we want to show that two notation systems describe the same segment (for instance $[0,3+\omega)$ and $[0,\omega($\;). For this purpose, one may prove that the two notation systems are order-isomorphic.

@@

{\it \textbf{Note:} This property reflects the equality $i+\omega=\omega$, that we will prove in larger notation systems, as well as in Schütte's model.}

@@

\textbf{Hint:}   Beware of the instances associated with sum and product of notations!
  You may consider additional fields such as ``there is a greatest element'' 
to make the sum and product of notations ``compositional''.

@@

Theorem embedding : 
     fun_bijection (nf: Ensemble T1)  (members epsilon0) inject.
 \end{Coqsrc}

\subsection{Remarks}
Let us recall that the library \href{../src/html/hydras.Schutte.Schutte.html}%
{\texttt{Schutte}} depends on five \emph{axioms} and lies explicitly in the  
framework of classical logic with a weak version of the axiom of choice
(please look at the documentation of
\href{https://coq.inria.fr/distrib/current/stdlib/Coq.Logic.ChoiceFacts.html}{\texttt{Coq.Logic.ChoiceFacts}}).
Nevertheless, the other modules:
\href{../src/html/hydras.Epsilon0.Epsilon0.html}%
{\texttt{Epsilon0}},
\href{../src/html/hydras.Hydra.Hydra.html}%
{\texttt{Hydra}}, et 
\href{../src/html/hydras.Gamma0.Gamma0.html}%
{\texttt{Gamma0}}
do not import any axioms and are really constructive.
